\vspace{-.35in} ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs \vspace{-.15in}

### 2023-10-27 coin-or/main.tex
coin-or and AIChE talk

\input{preamble}
\usepackage{wrapfig}
\title{ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
\author{Sungho Shin$^\dag$, François Pacaud$^\ddag$, and Mihai Anitescu$^\dag$}
\date{\small
  $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
}
\begin{document}
\maketitle

We are delighted to submit our nominations for the COIN-OR Cup,
highlighting our contributions of {\bf ExaModels} \cite{examodels} and
{\bf MadNLP} \cite{madnlp} in the field of computational
infrastructure for operations research, specifically in the area of
nonlinear programming (NLP). Our open-source software packages
represent a significant breakthrough in efficiently solving
large-scale NLPs by leveraging the capabilities of modern GPU
hardware.  In this document, we address the primary challenges and
opportunities for optimization software development in 2023,
particularly within the context of accelerated computing. We elucidate
how ExaModels and MadNLP collectively address technical obstacles and
capitalize on the advantages offered by modern accelerated
computing. Furthermore, we emphasize the remarkable performance
enhancements achieved in solving real-world nonlinear optimization
problems at scale. A noteworthy example of such progress is our recent
accomplishment in accelerating alternating current (AC) optimal power
flow (OPF) problems by up to a factor of ten compared to state-of-the-art
tools. Finally, we discuss how our development lays the foundation for
the advancement in various topics in operations research and unlocks
new possibilities.

\paragraph{Challenges and Opportunities.} In recent years, the
scalability of deep learning has experienced significant advancements,
primarily due to the utilization of NVIDIA
GPUs, which have demonstrated remarkable success in constructing
large-scale AI models. Additionally, in the high-performance computing (HPC) domain, GPU
accelerators have provided a substantial portion of the computational
power for next-generation leadership HPC systems such as Frontier and
Aurora, achieving the milestone of one exaflop per second.  While GPUs
have proven highly effective in various scientific computing areas
like machine learning and high-fidelity simulations, integrating them
into nonlinear optimization solvers has encountered
considerable challenges. Specifically, the seamless translation of
sparse automatic differentiation (AD) and sparse linear solver
routines to GPU architecture has been recognized to be particularly
challenging.  This is because conventional sparse AD and matrix
factorization algorithms are designed for CPUs and heavily rely on
serial computation, making them unsuitable for the single-instruction,
multiple-data (SIMD) parallelism employed by GPU architectures (see
Figure \ref{fig:simd}, left).

While GPU computation can accelerate several parts of the optimization
process straightforwardly, the slow data transfer between host and
device memory hinders the ad-hoc implementation of GPU accelerations
(Figure \ref{fig:simd}, right). To fully exploit the potential offered
by modern GPU hardware, it is crucial to develop a comprehensive
computational framework for optimization on GPUs. This requires
keeping both the problem data and the solver's intermediate computational
data exclusively within device memory, with the majority of
operations executed on the GPU. Consequently, the development of a
comprehensive NLP solution framework that performs
all necessary computations --- including AD, optimization, and linear
algebra --- exclusively on GPUs becomes imperative.  To address these
challenges, ExaModels and MadNLP offer a comprehensive solution
framework for NLP. They respectively enable AD and NLP solution
exclusively on GPUs, while linear algebra computation is performed by
external libraries (e.g., cuSOLVER), allowing for the full
utilization of GPUs.

\begin{figure}[t]
  \centering
  \scalebox{.82}{
    \begin{tikzpicture}[remember picture, scale=.9, font=\small]
    \node[draw,align=left] (I) at (0,2.75) {
      $\bullet$ Instruction $g(\cdot,\cdot)$\\
      $\bullet$ Pointer to array $x$
    };
    \node[draw] at (-4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (A\x) at (0,-\x/2) {$s_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (0,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (C\x) at (0,-\x/2) {$y_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \foreach\x in {1,2,3,4,5,7}{
      \draw[->] (A\x.east) -- (B\x.west);
      \draw[->] (B\x.east) -- (C\x.west);
      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
    }
    \node at (-4,-2.5) {Input Data};
    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
    \node at (4,-2.5) {Output Data};
  \end{tikzpicture}
  }
  \scalebox{.87}{
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

@@

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}
  \label{fig:simd}
\end{figure}

@@

Recently, we have discovered that the condensed-space interior point
method (IPM) with an inequality relaxation strategy is particularly
effective for solving large-scale NLPs, like AC optimal power flow, up
to moderate accuracy. This strategy efficiently addresses the
challenges posed by sparse matrix factorization on GPUs, as the
condensed KKT systems can be factorized without numerical pivoting,
which has previously hindered GPU utilization in the large-scale
optimization regime.  Furthermore, MadNLP offers diverse approaches
for handling KKT systems with special structures, via reduction
strategies \cite{pacaud2023accelerating} and Schur complement decomposition
strategies \cite{pacaud2023parallel}.

@@

Our numerical results, summarized in Table \ref{tab:num} (for detailed
results, please refer to \cite{shin2023accelerating}), clearly
demonstrate the significant potential of using GPUs. Specifically,
when applied to AC OPF problems, which are one of the most important
applications of NLP, we have achieved significant performance
enhancements. The GPU-based solutions achieve speedups of over 10
times compared to CPU-based state-of-the-art tools. Notably, ExaModels
enables derivative evaluations that are up to two orders of magnitude
faster, and the use of the condensed-space IPM strategy in MadNLP
greatly improves linear algebra computation speed. Furthermore, we
have also explored the utilization of these methods in reduced-space
OPF \cite{pacaud2023accelerating} and model predictive control (MPC)
\cite{cole2023exploiting}, where we have achieved similar degrees of
speedup.

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.
  \end{center}
  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
  \label{tab:num}
\end{table}

@@

\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA GPU & AMDGPU  & Intel GPU& Apple Metal\\
        \hline
        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\
        \hline
        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the NLP frameworks.}
  \label{tbl:portability}
\end{table}

@@

\begin{itemize}
\item \textit{Energy Infrastructures}: Enhancing the scalability of
optimization methods is crucial for improving the resilience of energy
system operations. We have demonstrated the ability of MadNLP
to solve large-scale security-constrained AC OPF
problems effectively with multiple GPUs~\cite{pacaud2023parallel}. This showcases the
potential of ExaModels and MadNLP to enhance the scalability of NLP by
utilizing GPU capabilities.

\item \textit{Model Predictive Control}: MPC has been widely used for
optimization-based control in various settings, such as chemical
processes, energy systems, and robotics. However, the effectiveness of
MPC is often limited by the latency of optimization solvers,
especially when short sampling times are required. Our results have
shown that MadNLP running on GPUs can solve such problems with high
efficiency \cite{cole2023exploiting}. Moving forward, we plan to
implement extension packages for control problems to enable
the control community to take advantage of accelerated NLP frameworks.

\item \textit{Machine Learning Surrogate Models}: An exciting future
direction is the integration of neural network surrogate models into
constrained optimization algorithms (a proof-of-concept presented in
\cite{shin2023constrained}). GPU acceleration plays a crucial role in
these applications, considering the complexity of real-world AI
models. MadNLP offers the necessary capabilities through the native
handling of device arrays as well as dense KKT systems with various
quasi-Newton strategies. We intend to explore these problems further
and provide extensions for surrogate modeling with GPU acceleration.

>>>>

\input{preamble}
\usepackage{wrapfig}
\title{
  \vspace{-.35in}
  ExaModels and MadNLP: An Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs
  \vspace{-.15in}
}
\author{
  Sungho Shin$^\dag$, François Pacaud$^\ddag$, and Mihai Anitescu$^\dag$
}
\date{\small
  \vspace{-.1in}
  $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
  \vspace{-.1in}
}
\begin{document}
\maketitle

We are delighted to submit our nominations for the COIN-OR Cup,
highlighting our contributions of {\bf ExaModels} \cite{examodels} and
{\bf MadNLP} \cite{madnlp} in the field of computational
infrastructure for operations research, specifically in the area of
nonlinear programming (NLP). Our open-source software packages
represent a significant breakthrough in efficiently solving
large-scale NLPs by leveraging the capabilities of modern GPUs.
In this document, we address the primary challenges and
opportunities for optimization software development in 2023,
particularly in the context of accelerated computing. We elucidate
how ExaModels and MadNLP address technical obstacles and
capitalize on the advantages offered by modern accelerated
computing. Furthermore, we emphasize the remarkable performance
enhancements achieved in solving real-world nonlinear optimization
problems at scale. A noteworthy example of such progress is our recent
accomplishment in accelerating alternating current (AC) optimal power
flow (OPF) problems by up to a factor of ten compared to state-of-the-art
tools. Finally, we discuss how our development lays the foundation for
the advancement in various topics in operations research and unlocks
new possibilities.

\paragraph{Challenges and Opportunities.} In recent years, the
scalability of deep learning has experienced significant advancements,
primarily due to the utilization of NVIDIA
GPUs, which have demonstrated remarkable success in constructing
large-scale AI models. Additionally, in the high-performance computing (HPC) domain, GPU
accelerators have provided a substantial portion of the computational
power for next-generation leadership HPC systems such as Frontier and
Aurora, achieving the milestone of one exaflop per second.  While GPUs
have proven highly effective in various scientific computing areas
like machine learning (ML) and simulations, integrating them
into nonlinear optimization solvers has encountered
considerable challenges. Specifically, the seamless translation of
sparse automatic differentiation (AD) and sparse linear solver
routines to GPU architecture has been recognized to be particularly
challenging.  This is because conventional sparse AD and matrix
factorization algorithms are designed for CPUs and heavily rely on
serial computation, making them unsuitable for the single-instruction,
multiple-data (SIMD) parallelism employed by GPU architectures (see
Figure \ref{fig:simd}, left).

While GPU computation can accelerate several parts of the optimization
process straightforwardly, the slow data transfer between host and
device memory hinders the ad-hoc implementation of GPU accelerations
(Figure \ref{fig:simd}, right). To fully exploit the potential offered
by modern GPU hardware, it is crucial to develop a comprehensive
computational framework for optimization on GPUs. This requires
keeping both the problem data and the solver's intermediate computational
data exclusively within device memory, with the majority of
operations executed on the GPU. Consequently, the development of a
comprehensive NLP solution framework that performs
all necessary computations---including AD, optimization, and linear
algebra---exclusively on GPUs becomes imperative.  To address these
challenges, ExaModels and MadNLP offer a comprehensive solution
framework for NLP. They respectively enable AD and NLP solution
exclusively on GPUs, while linear algebra computation is performed by
external libraries (e.g., cuSOLVER), allowing for the full
utilization of GPUs.

\begin{figure}[t]
  \centering
  \scalebox{.7}{
    \begin{tikzpicture}[remember picture, scale=.9, font=\small]
    \node[draw,align=left] (I) at (0,2.75) {
      $\bullet$ Instruction $g(\cdot,\cdot)$\\
      $\bullet$ Pointer to array $x$
    };
    \node[draw] at (-4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (A\x) at (0,-\x/2) {$s_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (0,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (C\x) at (0,-\x/2) {$y_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \foreach\x in {1,2,3,4,5,7}{
      \draw[->] (A\x.east) -- (B\x.west);
      \draw[->] (B\x.east) -- (C\x.west);
      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
    }
    \node at (-4,-2.5) {Input Data};
    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
    \node at (4,-2.5) {Output Data};
  \end{tikzpicture}
  }
  \scalebox{.75}{
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

@@

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
}
\vspace{-.075in}
\caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}
\vspace{-.075in}
\label{fig:simd}
\end{figure}

@@

Recently, we have discovered that the condensed-space interior point
method (IPM) with an inequality relaxation strategy is particularly
effective for solving large-scale NLPs, like AC optimal power flow, up
to moderate accuracy. This strategy efficiently addresses the
challenges posed by sparse matrix factorization on GPUs, as the
condensed KKT systems can be factorized without numerical pivoting,
which has previously hindered GPU utilization.  Furthermore, MadNLP
offers diverse approaches for handling KKT systems with special
structures, via reduction strategies \cite{pacaud2023accelerating} and
Schur complement decomposition strategies \cite{pacaud2023parallel}.

@@

Our numerical results for AC optimal power flow problems, summarized
in Table \ref{tab:num} (for detailed results, please refer to
\cite{shin2023accelerating}), clearly demonstrate the significant
potential of using GPUs. Specifically, when applied to AC OPF
problems, which are one of the most important applications of NLP, we
have achieved significant performance enhancements. The GPU-based
solutions achieve speedups of over 10 times compared to CPU-based
state-of-the-art tools. Notably, ExaModels enables derivative
evaluations that are up to two orders of magnitude faster, and the use
of the condensed-space IPM strategy in MadNLP greatly improves linear
algebra computation speed. Furthermore, we have also explored the
utilization of these methods in reduced-space OPF
\cite{pacaud2023accelerating} and model predictive control (MPC)
\cite{cole2023exploiting}, where we have achieved similar degrees of
speedup.

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
  \end{center}
  \vspace{-.075in}
  \caption{Solution times (sec) for solving AC OPF problems}
  \vspace{-.075in}
  \label{tab:num}
\end{table}

@@

\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA GPU & AMDGPU  & Intel GPU& Apple Metal\\
        \hline
        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\
        \hline
        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \vspace{-.075in}
  \caption{The current status on the portability of the NLP frameworks.}
  \vspace{-.075in}
  \label{tbl:portability}
\end{table}

@@

\begin{itemize}
\item \textit{Energy Infrastructures}: Enhancing the scalability of
optimization methods is crucial for improving the resilience of energy
system operations as the multi-period and stochastic formulations for
energy networks are large-scale. We have recently demonstrated the
ability of MadNLP to solve large-scale security-constrained AC OPF
problems effectively with multiple GPUs~\cite{pacaud2023parallel}.
This showcases the potential of accelerated computing, enabled by
ExaModels and MadNLP, to contribute to enhancing the grid efficiency
and resillience.

\item \textit{Model Predictive Control}: MPC has been widely used for
optimization-based control in various settings, such as chemical
processes, energy systems, and robotics. However, the effectiveness of
MPC is often limited by the latency of optimization solvers,
especially when short sampling times are required. Our results have
shown that MadNLP running on GPUs can solve such problems with high
efficiency \cite{cole2023exploiting}. Moving forward, we plan to
implement extension packages for control problems to enable
the control community to take advantage of better operations research
software with accelerated computing capabilities.

\item \textit{Machine Learning Surrogate Models}: An exciting future
direction is the integration of neural network surrogate models into
constrained optimization algorithms (proof-of-concept in
\cite{shin2023constrained}). The development of the new award-winning
software OMLT represents the increasing interest in incorporating ML
models into optimization problems \cite{ceccon2022omlt}. GPU
acceleration will play a crucial role in these applications,
considering the complexity of real-world AI models. MadNLP offers the
necessary capabilities through the native handling of device arrays as 
well as dense KKT systems with various quasi-Newton strategies. We
intend to explore these problems further and provide extensions for
surrogate modeling with GPU acceleration.
